Measuring the double-drift illusion with hand tracking: capturing illusion resets

### 2020-02-12 Manuscript.Rmd
convert manuscript to r markdown format

---
title: "Measuring the double-drift illusion with hand tracking: capturing illusion resets"
author: 
- Bernard Marius 't Hart 1,*
- Denise Y.P. Henriques 1
- Patrick Cavanagh 1,2,3
output:
  html_document:
    df_print: paged
  word_document: default
  pdf_document: default
bibliography: bibliography.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
# The functions here allow downloading the collected data from the OSF project:
source('R/data.R')
# This file has some common helper functions:
source('R/common.R')
# The functions in here are used on the data of the bounded tracking experiments:
source('R/tracking_bounded.R')
# And finally some code for the re-tracing experiment...
#source('R/retrace_onePass_V4.R')
# As well as models fit to that data:
source('R/models.R')
```

@@

If a gabor pattern drifts in one direction while its internal texture drifts in the orthogonal direction, its perceived direction deviates strongly from its true direction and is instead some combination of its real external motion and its internal motion (Tse & Hsieh, 2006). The perceived location of such a double-drift stimulus periodically resets toward the veridical position. There may indeed be a simple perdiodicity underlying these resets. However, since both the veridical and perceived location are available to the brain, these resets may also occur when these two positions are too far from each other to come from the same source stimulus. Here we test which of the two explanations is more likely. In a first experiment, we had participants continuously (for XX min) track the position of the perceived position of the stimulus. While the orientation of the trajectories reflect stimulus properties, there were no spontaneous nor any evoked resets.  That’s why, in a second experiment, we had participants re-trace the perceived trajectory of a short stimulus presentation, or indicate the initial direction of the stimulus they perceived. We found resets here, and they indicate a set spatial limit to the distance between the veridical and perceived location of the double-drift stimulus.

@@

Encoding the position of objects in the world is necessary for the myriad of visually guided motor tasks that we do every day. It is no surprise then that the brain has several mechanisms to gauge and update the position of objects in the world. For example, when a target is moving, its motion can influence it perceived location, making it appear to be slightly ahead of its true location (e.g. the flash lag, Nijhawan, 1994; or the flash grab effect, Cavanagh & Anstis, 2013). These cases of position extrapolation may be functional, helping the motor system overcome inevitable neural delays in targeting a moving object (Nijhawan, 1994; Duhamel et al., 1992). A stronger and very different motion-induced position shift arises when a moving gabor with internal motion orthogonal to its path is viewed in the periphery (Tse & Hsieh, 2006; Shapiro et al, 2010; Gurnsey & Baird, 2012; Kwon et al, 2015; Lisi & Cavanagh, 2015). This double-drift stimulus (Fig. 1, top left) generates extreme misjudgments of the moving target’s location which may deviate by as much as several degrees of visual angle from its true location. Surprisingly, this extraordinary perceptual illusion does not affect eye movements to the gabor: saccades to these targets are determined by their physical not their perceived location (Lisi & Cavanagh, 2015). This result indicates that both true and illusory locations are available in the brain. Indeed, although immediate eye-movements go towards the true position, delayed eye movements and pointing go to the perceived position [Lisi & Cavanagh, 2017; Massendari et al, 2018].

@@

The strength of the double-drift probably arises because of the very poor location information for a gabor pattern in the periphery when it has the same mean luminance as the background. With poor positional certainty, the motion information contributes to the location estimate, generating a shift in position. If the gabor itself is not moving (De Valois & De Valois, 1991), the shift saturates after about 100 ms (Chung et al., 2007) perhaps because the accumulating evidence of actual location of the stationary gabor patch is sufficient to anchor its position. However, when the gabor itself also moves as in the double-drift stimulus, there is no stable location information to anchor the position estimate and the build-up continues away from its physical location to reach deviations of several times the size of the gabor. At some point, the illusion resets, the perceived location jumps back to its physical location and then accumulation begins again (Fig. 1, bottom left). This long duration before resetting is well beyond the integration period of other visual effects and into the range of short-term memory. Here, we first examine the double-drift illusion by asking participants to trace out its path with a stylus on a tablet that is optically aligned with the monitor surface but with the hand not visible. Earlier results for pointing to the double-drift target (Lisi & Cavanagh, 2017) show that the illusion persists for manual pointing. However, with online tracking, the presence of the hand near the location of the perceived gabor may interact with the illusion.

@@

@@

For the first task 4 participants were recruited from the lab (3 female; ages 21 – 31, mean: 25.75). For the second task,  6 participants were recruited from the lab (4 female; ages 19 – 26, mean: 21.6). All participants reported to be right handed and have either normal or corrected to normal vision.  Procedures were approved by York’s Human Participants Review Committee, and all participants provided prior, written, informed consent.

## Setup
Participants used a stylus on a horizontal drawing tablet (Intuos Pro) to indicate where they perceived the location of a gabor (see Fig 1e). Stimuli were displayed on a downward facing LED monitor (30 Hz, XX”, 1440x900), level with the drawing tablet. These were observed through an upward facing mirror placed exactly between the tablet and monitor, so that the stimuli appear to be in the same plane as the hand.
- size of stimuli in cm, speed in cm/s (with guesstimates of dva?)

@@

All analyses were done in R 3.6.1 [@Rproject]

@@

- 9 (exp 1; x female; age: y-z) + 9 (exp 2; x female; age y-z)
- Informed consent, ethics board

The same hardware was used as in the continuous tracking experiment, except that in one of the tasks, participants could now respond using the arrow keys on a keypad.

@@

Participants alternated between blocks of two kinds of trials (in counterbalanced order). In both types of trial, the gabor was shown starting close to the participant and moving away in a forward direction, for a single pass along the workspace, taking 4 seconds. During presentation of the gabor, the participant was instructed to maintain fixation to the left or right of the stimulus. After the gabor disappeared, the participant could break fixation, and was to report the percept in one of two ways. In one kind of block, the participant re-traced the perceived path of the gabor using the stylus. The drawn path was shown as a line on the screen. Participants could “key in” their re-traced percept by pressing ENTER or start over by pressing the ESCAPE key. In the other kind of block, participants reported the initial direction of the gabor (before any resets occurred), by using the LEFT and RIGHT arrow keys to change the orientation of a line of XX cm, with it’s origin in the same spot as where the gabor started moving. Participants could change the line’s orientation it until satisfied and then “key in” their response by pressing ENTER. The arrow report task has been used before (ref?) and can be used as validation of this task. The re-tracing task has also been used before, but in order to test evoked resets (Nakayama & Holcombe, 2019).

>>>>

---
title: "Measuring the double-drift illusion with hand tracking: capturing illusion resets"
author: 
- Bernard Marius 't Hart 1,*
- Denise Y.P. Henriques 1
- Patrick Cavanagh 1,2,3
output:
  html_document:
    df_print: paged
  pdf_document: default
  word_document: default
bibliography: bibliography.bib
csl: journal-of-vision.csl
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
# The functions here allow downloading the collected data from the OSF project:
source('R/data.R')
# This file has some common helper functions:
source('R/common.R')
# The functions in here are used on the data of the bounded tracking experiments:
source('R/tracking_bounded.R')
# And finally some code for the re-tracing experiment...
source('R/retrace_onePass_V4.R')
# As well as models fit to that data:
source('R/models.R')
```

@@

If a gabor pattern drifts in one direction while its internal texture drifts in the orthogonal direction, its perceived direction deviates strongly from its true direction and is instead some combination of its real external motion and its internal motion [@Tse2006]. The perceived location of such a double-drift stimulus periodically resets toward the veridical position. There may indeed be a simple perdiodicity underlying these resets. However, since both the veridical and perceived location are available to the brain, these resets may also occur when these two positions are too far from each other to come from the same source stimulus. Here we test which of the two explanations is more likely. In a first experiment, we had participants continuously (for XX min) track the position of the perceived position of the stimulus. While the orientation of the trajectories reflect stimulus properties, there were no spontaneous nor any evoked resets.  That’s why, in a second experiment, we had participants re-trace the perceived trajectory of a short stimulus presentation, or indicate the initial direction of the stimulus they perceived. We found resets here, and they indicate a set spatial limit to the distance between the veridical and perceived location of the double-drift stimulus.

@@

Encoding the position of objects in the world is necessary for the myriad of visually guided motor tasks that we do every day. It is no surprise then that the brain has several mechanisms to gauge and update the position of objects in the world. For example, when a target is moving, its motion can influence it perceived location, making it appear to be slightly ahead of its true location (e.g. the flash lag, @Nijhawan1994; or the flash grab effect, @Cavanagh2013). These cases of position extrapolation may be functional, helping the motor system overcome inevitable neural delays in targeting a moving object [@Nijhawan1994; @Duhamel1992]. A stronger and very different motion-induced position shift arises when a moving gabor with internal motion orthogonal to its path is viewed in the periphery [@Tse2006; @Shapiro2010; @Gurnsey2012; @Kwon2015; @Lisi2015]. This double-drift stimulus (Fig. 1, top left) generates extreme misjudgments of the moving target’s location which may deviate by as much as several degrees of visual angle from its true location. Surprisingly, this extraordinary perceptual illusion does not affect eye movements to the gabor: saccades to these targets are determined by their physical not their perceived location [@Lisi2015]. This result indicates that both true and illusory locations are available in the brain. Indeed, although immediate eye-movements go towards the true position, delayed eye movements and pointing go to the perceived position [@Lisi2017; @Massendari2018].

@@

The strength of the double-drift probably arises because of the very poor location information for a gabor pattern in the periphery when it has the same mean luminance as the background. With poor positional certainty, the motion information contributes to the location estimate, generating a shift in position. If the gabor itself is not moving [@DeValois1991], the shift saturates after about 100 ms [@Chung2007], perhaps because the accumulating evidence of actual location of the stationary gabor patch is sufficient to anchor its position. However, when the gabor itself also moves as in the double-drift stimulus, there is no stable location information to anchor the position estimate and the build-up continues away from its physical location to reach deviations of several times the size of the gabor. At some point, the illusion resets, the perceived location jumps back to its physical location and then accumulation begins again (Fig. 1, bottom left). This long duration before resetting is well beyond the integration period of other visual effects and into the range of short-term memory. Here, we first examine the double-drift illusion by asking participants to trace out its path with a stylus on a tablet that is optically aligned with the monitor surface but with the hand not visible. Earlier results for pointing to the double-drift target [@Lisi2017] show that the illusion persists for manual pointing. However, with online tracking, the presence of the hand near the location of the perceived gabor may interact with the illusion.

@@

# Methods

For the first task in experiment 1, 4 participants were recruited from the lab (3 female; ages 21 – 31, mean: 25.75). For the second task in experiment 1, 6 participants were recruited from the lab (4 female; ages 19 – 26, mean: 21.6). For experiment 2, 9 participants were recruited from the lab (6 female; ages 19 - 27, mean: 22.8). All participants reported to be right handed and have either normal or corrected to normal vision.  Procedures were approved by York’s Human Participants Review Committee, and all participants provided prior, written, informed consent.

Participants used a stylus on a horizontal drawing tablet (Intuos Pro) to indicate where they perceived the location of a gabor (see Fig 1e), as well as a keypad for additional responses. Stimuli were displayed on a downward facing LED monitor (30 Hz, 20”, 1680x1050), level with the drawing tablet. These were observed through an upward facing mirror placed exactly between the tablet and monitor, so that the stimuli appear to be in the same horizontal plane as the hand. The stimuli were phase-changing gabor patches 
- size of stimuli in cm, speed in cm/s (with guesstimates of dva?)

Experiments were run in Python 2.7 with PsychoPy [@Peirce2019] and analyses were done in R 3.6.1 [@Rproject].

@@

@@

@@

@@

Participants alternated between blocks of two kinds of trials (in counterbalanced order). In both types of trial, the gabor was shown starting close to the participant and moving away in a forward direction, for a single pass along the workspace, taking 4 seconds. During presentation of the gabor, the participant was instructed to maintain fixation to the left or right of the stimulus. After the gabor disappeared, the participant could break fixation, and was to report the percept in one of two ways. In one kind of block, the participant re-traced the perceived path of the gabor using the stylus. The drawn path was shown as a line on the screen. Participants could “key in” their re-traced percept by pressing ENTER or start over by pressing the ESCAPE key. In the other kind of block, participants reported the initial direction of the gabor (before any resets occurred), by using the LEFT and RIGHT arrow keys to change the orientation of a line of XX cm, with it’s origin in the same spot as where the gabor started moving. Participants could change the line’s orientation it until satisfied and then “key in” their response by pressing ENTER. The arrow report task has been used before (ref?) and can be used as validation of this task. The re-tracing task has also been used before, but in order to test evoked resets [@Nakayama2019].
